---
title: "PA1_template.Rmd"
author: "Umar Faruk"
date: "12/19/2020"
output:
  html_document: default
  pdf_document: default
---




#Introduction
It is now possible to collect a large amount of data about personal movement using activity monitoring devices such as a [Fitbit](http://www.fitbit.com/), [Nike Fuelband](http://www.nike.com/us/en_us/c/nikeplus-fuelband), or [Jawbone Up](https://jawbone.com/up). These type of devices are part of the “quantified self” movement – a group of enthusiasts who take measurements about themselves regularly to improve their health, to find patterns in their behavior, or because they are tech geeks. But these data remain under-utilized both because the raw data are hard to obtain and there is a lack of statistical methods and software for processing and interpreting the data.

This assignment makes use of data from a personal activity monitoring device. This device collects data at 5 minute intervals through out the day. The data consists of two months of data from an anonymous individual collected during the months of October and November, 2012 and include the number of steps taken in 5 minute intervals each day.

The data for this assignment can be downloaded from the course web site:

Dataset: [Activity monitoring data](https://d396qusza40orc.cloudfront.net/repdata%2Fdata%2Factivity.zip)
The variables included in this dataset are:

*steps:* Number of steps taking in a 5-minute interval (missing values are coded as 𝙽𝙰)
*date:* The date on which the measurement was taken in YYYY-MM-DD format
*interval:* Identifier for the 5-minute interval in which measurement was taken
The dataset is stored in a comma-separated-value (CSV) file and there are a total of 17,568 observations in this dataset.

#Loading and preprocessing the data

 Activation of packages and unziping data to get csv file
``` {r}
library("data.table")
library(ggplot2)

fileUrl <- "http://d396qusza40orc.cloudfront.net/repdata%2Fdata%2Factivity.zip"
download.file(fileUrl, destfile = paste0(getwd(),'/repdata%2Fdata%2Factivity.zip'), method = "curl")
unzip("repdata%2Fdata%2Factivity.zip", exdir = "data")

activityDT<- data.table::fread(input = "data/activity.csv")
activityDT
summary(activityDT)
```


Reading csv data
```{r}
Total_Steps <- activityDT[, c(lapply(.SD, sum, na.rm= FALSE)), .SDcols = c("steps"), by= .(date)]
head(Total_Steps, 10)
```

Plot
``` {r}
ggplot(Total_Steps, aes(x= steps)) + 
  geom_histogram(fill="blue", binwidth = 1000)+ labs(title = "Daily Steps", x= "steps", y= "Frequency")
```


Calculation and the reporting of the mean and median of the total number of steps taken per day
``` {r}
Total_Steps[, .(Mean_Steps= mean(steps, na.rm = TRUE), Median_Steps = median(steps, na.rm = TRUE))]
```
 



Make a time series plot (i.e. 𝚝𝚢𝚙𝚎 = "𝚕") of the 5-minute interval (x-axis) and the average number of steps taken, averaged across all days (y-axis)

```{r}
IntervalDT <- activityDT[, c(lapply(.SD, mean, na.rm = TRUE)), .SDcols = c("steps"), by = .(interval)]
ggplot(IntervalDT, aes(x= interval, y=steps)) + geom_line(color= "blue", size=1) + labs(title = "Avg. Daily Steps", x= "Interval", y = "Avg. Steps per day")
```

  Which 5-minute interval, on average across all the days in the dataset, contains the maximum number of steps?
```{r}
IntervalDT[steps == max(steps), .(max_interval = interval)]
```


  Calculate and report the total number of missing values in the dataset
```{r}
activityDT[is.na(steps), .N ]
```


##Devise a strategy for filling in all of the missing values in the dataset. The strategy does not need to be sophiscated. For example, you could use the mean/median for that day, or the mean for the 5-minute interval etc.

Filling in missing values with median of dataset.
    
    ``` {r}
activityDT[is.na(steps), "steps"]<- activityDT[, c(lapply(.SD, median, na.rm = TRUE)), 
                                               .SDcols = c("steps")]
    ```

  Create a new dataset that is equal to the original dataset but with the missing data filled in.

```{r}
data.table::fwrite(x = activityDT, file= "data/tidyData.csv", quote = FALSE)
```

Make a histogram of the total number of steps taken each day and calculate and report the mean and median total number of steps taken per day. Do these values differ from the estimates from the first part of the assignment? What is the impact of imputing missing data on the estimates of the total daily number of steps? 

total number of steps taken per day

```{r}
Total_Steps<- activityDT[, c(lapply(.SD, sum)), .SDcols = c("steps"), by = .(date)]
```

mean and median total number of steps taken per day

```{r}
Total_Steps[, .(Mean_Steps = mean(steps), Median_Steps = median(steps))]
```



```{r}
ggplot(Total_Steps, aes(x=steps)) + geom_histogram(fill = "blue", binwidth = 1000) + labs(title= "Daily Steps", x= "Steps", y= "Frequency")
```

#Are there differences in activity patterns between weekdays and weekends?

Just recreating activityDT from scratch then making the new factor variable. (No need to, just want to be clear on what the entire process is.) 

``` {r}
activityDT <- data.table::fread(input = "data/activity.csv")
activityDT[, date := as.POSIXct(date, format = "%Y-%m-%d")]
activityDT[, `Day of Week`:= weekdays(x = date)]
activityDT[grepl(pattern = "Monday|Tuesday|Wednesday|Thursday|Friday", x = `Day of Week`), "weekday or weekend"] <- "weekday"
activityDT[grepl(pattern = "Saturday|Sunday", x = `Day of Week`), "weekday or weekend"] <- "weekend"
activityDT[, `weekday or weekend` := as.factor(`weekday or weekend`)]
head(activityDT, 10)
```

Make a panel plot containing a time series plot (i.e. 𝚝𝚢𝚙𝚎 = "𝚕") of the 5-minute interval (x-axis) and the average number of steps taken, averaged across all weekday days or weekend days (y-axis). See the README file in the GitHub repository to see an example of what this plot should look like using simulated data.

Devise a strategy for filling in all of the missing values in the dataset. The strategy does not need to be sophisticated. For example, you could use the mean/median for that day, or the mean for that 5-minute interval, etc.

``` {r}
activityDT[is.na(steps), "steps"] <- activityDT[, c(lapply(.SD, median, na.rm = TRUE)), .SDcols = c("steps")]
IntervalDT <- activityDT[, c(lapply(.SD, mean, na.rm = TRUE)), .SDcols = c("steps"), by = .(interval, `weekday or weekend`)]
```

Plotting of average daily steps by weektype

``` {r}
ggplot(IntervalDT , aes(x = interval , y = steps, color=`weekday or weekend`)) + geom_line() + labs(title = "Avg. Daily Steps by Weektype", x = "Interval", y = "No. of Steps") + facet_wrap(~`weekday or weekend` , ncol = 1, nrow=2)
```